---
title: "Job Training vs. Real Earnings"
format: html
author: Winnie Hung
---

### Imports and Libraries

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("wooldridge")
library(wooldridge)
install.packages('patchwork')
library(patchwork)
```

### Pull Dataset

```{r}
data(jtrain2, package='wooldridge')
head(jtrain2)
```

### Variables Description

```{r}
?jtrain2
```

### Exploratory Data Analysis

```{r}
p_1 <- ggplot(jtrain2, aes(x = re78)) + 
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(title = "Histogram of re78", x = "Real Earnings in 1978 ($1000s)", y = "Count") +
  theme(axis.title.x = element_text(angle = 30, hjust = 1))

p_2 <- ggplot(jtrain2, aes(x = re75)) +
  geom_histogram(binwidth = 2, fill = "tomato", color = "white") +
  labs(title = "Histogram of re75", x = "Real Earnings in 1975 ($1000s)", y = "Count") +
  theme(axis.title.x = element_text(angle = 30, hjust = 1))

p_3 <- ggplot(jtrain2, aes(x = re74)) +
  geom_histogram(binwidth = 2, fill = "darkgreen", color = "white") +
  labs(title = "Histogram of re74", x = "Real Earnings in 1974 ($1000s)", y = "Count") +
  theme(axis.title.x = element_text(angle = 30, hjust = 1))

p_1 + p_2 + p_3
```

```{r}
p_4 <- ggplot(jtrain2, aes(x = lre78)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(title = "Histogram of re78", x = "Log Real Earnings in 1978 ($1000s)", y = "Count") +
  theme(axis.title.x = element_text(angle = 30, hjust = 1))

p_5 <- ggplot(jtrain2, aes(x = lre75)) +
  geom_histogram(binwidth = 2, fill = "tomato", color = "white") +
  labs(title = "Histogram of re75", x = "Log Real Earnings in 1975 ($1000s)", y = "Count") +
  theme(axis.title.x = element_text(angle = 30, hjust = 1))

p_6 <- ggplot(jtrain2, aes(x = lre74)) +
  geom_histogram(binwidth = 2, fill = "darkgreen", color = "white") +
  labs(title = "Histogram of re74", x = "Log Real Earnings in 1974 ($1000s)", y = "Count") +
  theme(axis.title.x = element_text(angle = 30, hjust = 1))

p_4 + p_5 + p_6
```

```{r}
p_7 <- ggplot(jtrain2, aes(x = age)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(title = "Histogram of age", x = "Log Real Earnings in 1978 ($1000s)", y = "Count")

p_8 <- ggplot(jtrain2, aes(x = mostrn)) +
  geom_histogram(binwidth = 2, fill = "tomato", color = "white") +
  labs(title = "Histogram of months in training", x = "Log Real Earnings in 1975 ($1000s)", y = 
         "Count")

p_9 <- ggplot(jtrain2, aes(x = educ)) +
  geom_histogram(binwidth = 2, fill = "darkgreen", color = "white") +
  labs(title = "Histogram of education", x = "Log Real Earnings in 1974 ($1000s)", y = "Count")

p_10 <- ggplot(jtrain2, aes(x = factor(unem78), fill = factor(black))) + geom_bar(position =
                                                                                    "fill")

p_7 + p_8 + p_9 + p_10
```

### Models

```{r}
lr_2 <- lm(unem78 ~ black + educ + re74 + re75 + train, data=jtrain2)
summary(lr_2)
```

-   black = 0.2077: holding educ, re74, re75, train fixed, being Black is associated with about a +0.208 higher probability of unem78=1.

-   train = -0.1122: being assigned to training is associated with about a -0.112 lower probability of unem78=1.

-   Model isn't explaining much variation in umem78 even though black nd train are statistically significant.

### More Visualization

```{r}
ggplot(data = jtrain2, mapping = aes(x = factor(train), y = re78)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  labs(x = "Job Training (0 = No, 1 = Yes)", y = "Log Real Earnings
       in 1978")
```

```{r}
table(jtrain2$train, jtrain2$black)

boxplot(jtrain2$educ ~ jtrain2$train)
```

#### Correlation Test on Variables in this Topic. 

```{r}
jtrain2 |> 
  cor(jtrain2[c("train", "educ", "re74", "re75")])
```

-   Job Training is positively correlated (0.1335) with 1978 earnings.

-   But the correlation is small(\`0.10-0.13).

-   Regression is more meaningful than correlation.

#### More Modeling

Trying to test model performance with different variables:

-   feature engineering might help.

```{r}
lin_reg1 <- lm(re78 ~ train + educ + black + re74 + re75, data
               =jtrain2)
summary(lin_reg1)
```

-   Individuals assigned to training earned \$1,711 more in 1978, on average, holding education, race, and prior earnings constant.

    -   Job Training does increase real earnings with about \$1700 positive effect.

-   Black participants earned about \$2,160 less in 1978, on average, conditional on education and prior earnings.

-   Each additional year of education is associated with approximately \$405 higher earnings in 1978, holding other factors constant.

-   Prior earnings are included to control for baseline labor market attachment. Although their coefficients are positive, they are not statistically significant int his specification.

-   The model explains about 5% if the variation in 1978 earnings, which is typical in individual-level earnings regressions.

```{r}
lm_reg2 <- lm(lre78 ~ train + educ + black + lre74 + lre75, data
              =jtrain2)
summary(lm_reg2)
```

#### Effect of Training

-   train = 0.23352

-   p = 0.029985

-   Percent effect \~= 100 \* 0.2335 = 23.35%

    -   e\^(0.2335) - 1 = 26.3%

-   Participation in the training program increased 1978 earnings by approximately 23-26 percent, holding education, race, and prior earnings constant. This effect is statistically significant at the 5% level.

#### Black Coefficient

-   black = -0.57754

-   p \< 0.001

    -   e\^(-0.5775) - 1 \~= - 44%

-   Black participants earned roughly 40-45 percent less than non-Black participants, conditional on other covariates.

#### Education

-   educ = 0.0358

-   p = 0.228

-   each additional year of education \~= 3.6% higher earnings, but not statistically significant.

### Summary

-   The log model is has a better performance because it shows a large percentage effect and this is randomize data, it has causal interpretation.

-   The log-earnings regression indicates that participation in the job training program increased 1978 earnings by approximately 23-26 percent, holding education, race, and prior earnings constant. This effect is statistically significant at the 5 percent level. black participants earned roughly 40-45 percent less than non-Black participants, conditional on covariates. Prior earnings from 1975 are marginally predictive of 1978 earnings, while 1974 earnings are not statistically significant. Overall, the results suggest that the training program has a sizable and economically meaningful impact on earnings.

### Is There a Omitted Variable Bias (OVB) in the Model?

Is OVB if:

1.  It affects the dependent variable (which is earnings)
2.  It is correlated with the train variable

There is not OVB if:

1.  either condition above fails –\> that means there are no bias in the train coefficient.

```{r}
lm(lre78 ~ train, data=jtrain2)

lm(lre78 ~ train + educ + black + lre74 + lre75, data=jtrain2)

lm(lre78 ~ train + educ + black + hisp + age + agesq + 
     married + nodegree + 
     lre74 + lre75 + unem74 + unem75, 
   data=jtrain2)

lm(train ~ educ + black + age + married + nodegree + lre74 + lre75, data=jtrain2)
```

-   There is little evidence of OVB affecting the training effect because the train coefficient decreases slightly, and it does not collapse, flip sign, or dramatically change.

-   The coefficient on training remains stable, decreasing slightly from 0.245 in the simple model to 0.208 in the fully controlled model. This stability suggest that omitted variables are unlikely to substantially bias the estimated treatment effect. Additionally, a regression of training status on observable characteristics shows weak correlations, consistent with random assignment. Therefore, the estimated training effect appears robust and plausibly causal.

#### Result

-   With no OVB in the model, the training effect remains around 20-25% increase in earnings.
